### 2025-07-09 manuscript/part2.tex
type system axiomatic

\subsection*{Base axioms}
\[
\begin{array}{rcll}
  \tau & ::= & \mathsf{Bool} 
            \;\bigm|\; \mathsf{Int}_{s,w}
            \;\bigm|\; \mathsf{Array}<\mathsf{Int}_{s,w}>
  & (s\in\{\top,\bot\},\;0<w<2^{29}-1) \\[1ex]
  \Gamma & ::= & \emptyset 
               \;\bigm|\; \Gamma,\,x:\tau
\end{array}
\]
\\
\[
  \frac{}{\Gamma \vdash \mathit{true} : \mathsf{Bool}}
  \; (\textsc{T-True}), \quad
  \frac{}{\Gamma \vdash \mathit{false} : \mathsf{Bool}}
  \; (\textsc{T-False})
\]
\[
  \frac{}
  {\Gamma \;\vdash\; n^{(s,w)} : \mathsf{Int}_{s,w}}
  \; (\textsc{T-IntLit}), \quad
  where\;
  s \in \{\top, \bot\},
  \quad
  w \in \mathbb{N},
  \quad
  1 \;\le\; w \;\le\; 2^{29}-1
\]

\subsection*{Inference rules}

\[
  \begin{aligned}
    \frac{x:\tau\in\Gamma}{\Gamma \vdash x : \tau}
     & \quad(\textsc{T-Var})   \\[2ex]
    \frac{\Gamma \vdash e : \tau}{\Gamma \vdash (\,e\,) : \tau}
     & \quad(\textsc{T-Paren})
  \end{aligned}
\]
\[
  \frac{\Gamma \vdash e : \mathsf{Bool}}
  {\Gamma \vdash !\,e : \mathsf{Bool}}
  \quad(\textsc{T-Not})
\]
\[
  \begin{aligned}
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash e_1 * e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,w_{\,\max(w_1, w_2)}}}
     & \quad(\textsc{T-Mult}) \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash e_1 \,/\, e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,w_{\,\max(w_1, w_2)}}}
     & \quad(\textsc{T-Div})  \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash e_1 \% e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,w_{\,\min(w_1, w_2)}}}
     & \quad(\textsc{T-Mod})  \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash e_1 + e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,w_{\,\max(w_1, w_2)}}}
     & \quad(\textsc{T-Add})  \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash e_1 - e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,w_{\,\max(w_1, w_2)}}}
     & \quad(\textsc{T-Sub})
  \end{aligned}
\]

\[
  \begin{aligned}
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
    {\Gamma \vdash e_1 < e_2   : \mathsf{Bool}}
     & (\textsc{T-Lt}) \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
    {\Gamma \vdash e_1 \le e_2 : \mathsf{Bool}}
     & (\textsc{T-Le}) \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
    {\Gamma \vdash e_1 > e_2   : \mathsf{Bool}}
     & (\textsc{T-Gt}) \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
    {\Gamma \vdash e_1 \ge e_2 : \mathsf{Bool}}
     & (\textsc{T-Ge})
  \end{aligned}
\]
\[
  \begin{aligned}
    \frac{\Gamma \vdash e_1 : \tau
      \quad
      \Gamma \vdash e_2 : \tau}
    {\Gamma \vdash e_1 == e_2 : \mathsf{Bool}}
     & \quad(\textsc{T-Eq})  \\[1ex]
    \frac{\Gamma \vdash e_1 : \tau
      \quad
      \Gamma \vdash e_2 : \tau}
    {\Gamma \vdash e_1 \:!= e_2 : \mathsf{Bool}}
     & \quad(\textsc{T-Neq})
  \end{aligned}
\]
\[
  \begin{aligned}
    \frac{\Gamma \vdash e_1 : \mathsf{Bool}
      \quad
      \Gamma \vdash e_2 : \mathsf{Bool}}
    {\Gamma \vdash e_1 \;\code{ &                  & }\; e_2 : \mathsf{Bool}}
                                & (\textsc{T-And})                            \\[1ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Bool}
      \quad
      \Gamma \vdash e_2 : \mathsf{Bool}}
    {\Gamma \vdash e_1 \;\texttt{||}\; e_2 : \mathsf{Bool}}
                                & (\textsc{T-Or})
  \end{aligned}
\]
\[
  \frac{\Gamma \vdash e_1 : \tau
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_3 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_4 : \tau
  }
  {\Gamma \vdash e_1 \:\{e_2 \::\: e_3\}\: e_4 : \tau} \quad(\textsc{T-ProbabilisticAssign.})
\]
\[
  \begin{aligned}
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s, w}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
    {\Gamma \vdash \mathrm{Bernoulli}(e_1,e_2) : \mathsf{Int}_{\bot,1}}
     & \quad(\textsc{T-Bernoulli}) \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Uniform}(e_1,e_2) : \mathsf{Int}_{s_2,w_2}}
     & \quad(\textsc{T-Uniform})
  \end{aligned}
\]

\subsection*{Arrays in the Type System}

\paragraph{Type‐Formation}
\[
  \frac{\vdash\;\tau\;\mathsf{type}}
       {\vdash\;\tau\;\mathsf{Array}\;\mathsf{type}}
  \quad(\textsc{Ty-Array})
\]

\paragraph{Array Literals (immutable)}  
\[
  \frac{\forall i\in 1..n.\;\Gamma\vdash e_i:\tau}
       {\Gamma\vdash [\,e_1,e_2,\dots,e_n\,] : \tau\;\mathsf{Array}}
  \quad(\textsc{T-ArrayLit})
\]

\paragraph{Indexing}
\[
  \frac{\Gamma\vdash e : \tau\;\mathsf{Array}
        \quad
        \Gamma\vdash i : \mathsf{Int}_{s,w}}
       {\Gamma\vdash e[i] : \tau}
  \quad(\textsc{T-Index})
\]

\paragraph{Update (mutable arrays)}  
\[
  \frac{\Gamma\vdash e_1 : \tau\;\mathsf{Array}
        \quad
        \Gamma\vdash i : \mathsf{Int}_{s,w}
        \quad
        \Gamma\vdash e_2 : \tau}
       {\Gamma\vdash e_1[i] = e_2 : \mathsf{Unit}}
  \quad(\textsc{T-Store})
\]

\paragraph{Length (optional)}  
\[
  \frac{\Gamma\vdash e : \tau\;\mathsf{Array}}
       {\Gamma\vdash \mathsf{length}(e) : \mathsf{Int}_{\bot,\,w_{\mathit{len}}}}
  \quad(\textsc{T-Length})
\]

\subsection{Worked Example}
To derive the judgment

$$
  \emptyset \vdash \mathtt{if}\;\mathtt{true}\;\mathtt{then}\;(1+2)\;\mathtt{else}\;3 : \mathsf{Int},
$$

we construct the following proof:

$$
  \dfrac{
    \dfrac{}{\emptyset \vdash \mathtt{true} : \mathsf{Bool}} \; (\textsc{T-True})
    \quad
    \dfrac{
      \dfrac{}{\emptyset \vdash 1 : \mathsf{Int}} \quad
      \dfrac{}{\emptyset \vdash 2 : \mathsf{Int}}
    }{\emptyset \vdash 1+2 : \mathsf{Int}} \; (\textsc{T-Plus})
    \quad
    \dfrac{}{\emptyset \vdash 3 : \mathsf{Int}} \; (\textsc{T-Int})
  }{
    \emptyset \vdash \mathtt{if}\;\mathtt{true}\;\mathtt{then}\;(1+2)\;\mathtt{else}\;3 : \mathsf{Int}
  } \; (\textsc{T-If})
$$

\begin{align*}
   & \textbf{T-Assign}
   &                   & \dfrac{Γ ⊢ e : τ \quad x:τ ∈ Γ}
  {Γ ⊢ x = e\;\mathsf{ok}}                                     \\[1ex]
   & \textbf{T-IfSt}
   &                   & \dfrac{Γ ⊢ e : \mathsf{Bool}
    \quad Γ ⊢ S_1\;\mathsf{ok}
    \quad Γ ⊢ S_2\;\mathsf{ok}}
  {Γ ⊢ \mathtt{if}\,(e)\,S_1\,\mathtt{else}\,S_2\;\mathsf{ok}} \\[1ex]
   & \textbf{T-Seq}
   &                   & \dfrac{Γ ⊢ S_1\;\mathsf{ok}
    \quad Γ ⊢ S_2\;\mathsf{ok}}
  {Γ ⊢ S_1; S_2\;\mathsf{ok}}
\end{align*}

>>>>

\subsection*{Base axioms}
\[
  \begin{array}{rcll}
    \tau   & ::=                                & \mathsf{Bool}
    \;\bigm|\; \mathsf{Int}_{s,w}
    \;\bigm|\; \mathsf{Array}<\mathsf{Int}_{s,w}>
           & (s\in\{\top,\bot\},\;0<w<2^{29}-1)                 \\[1ex]
    \Gamma & ::=                                & \emptyset
    \;\bigm|\; \Gamma,\,x:\tau
  \end{array}
\]
\\
\[
  \frac{}{\Gamma \vdash \mathit{true} : \mathsf{Bool}}
  \; (\textsc{T-True}), \quad
  \frac{}{\Gamma \vdash \mathit{false} : \mathsf{Bool}}
  \; (\textsc{T-False})
\]
\[
  \frac{}
  {\Gamma \;\vdash\; n^{(s,w)} : \mathsf{Int}_{s,w}}
  \; (\textsc{T-IntLit}), \quad
  where\;
  s \in \{\top, \bot\},
  \quad
  w \in \mathbb{N},
  \quad
  1 \;\le\; w \;\le\; 2^{29}-1
\]
\[
  \frac{\vdash\;\mathsf{Int}_{s,w}\;}
  {\vdash\;\mathsf{Array}<\mathsf{Int}_{s,w}>}
  \quad(\textsc{Ty-Array})
\]

\subsection*{Inference rules}
\begin{multicols}{2}

  %–– Column 1 ––%
  \[
    \begin{aligned}
      \frac{x:\tau\in\Gamma}{\Gamma \vdash x : \tau}
       & (\textsc{T-Var})   \\[2ex]
      \frac{\Gamma \vdash e : \tau}{\Gamma \vdash (\,e\,) : \tau}
       & (\textsc{T-Paren}) \\[2ex]
      \frac{\Gamma \vdash e : \mathsf{Bool}}
      {\Gamma \vdash \neg e : \mathsf{Bool}}
       & (\textsc{T-Not})
    \end{aligned}
  \]

  \[
    \begin{aligned}
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
      {\Gamma \vdash e_1 * e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,\max(w_1, w_2)}}
       & (\textsc{T-Mult}) \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
      {\Gamma \vdash e_1 / e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,\max(w_1, w_2)}}
       & (\textsc{T-Div})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
      {\Gamma \vdash e_1 \bmod e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,\min(w_1, w_2)}}
       & (\textsc{T-Mod})  \\[2ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
      {\Gamma \vdash e_1 + e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,\max(w_1, w_2)}}
       & (\textsc{T-Add})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
      {\Gamma \vdash e_1 - e_2 : \mathsf{Int}_{s_{\,s_1\lor s_2},\,\max(w_1, w_2)}}
       & (\textsc{T-Sub})
    \end{aligned}
  \]

  %–– Column 2 ––%
  \[
    \begin{aligned}
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 < e_2   : \mathsf{Bool}}
       & (\textsc{T-Lt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 <= e_2 : \mathsf{Bool}}
       & (\textsc{T-Le})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 > e_2   : \mathsf{Bool}}
       & (\textsc{T-Gt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 >= e_2 : \mathsf{Bool}}
       & (\textsc{T-Ge})  \\[2ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 == e_2 : \mathsf{Bool}}
       & (\textsc{T-Eq})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 \:!= e_2 : \mathsf{Bool}}
       & (\textsc{T-Neq}) \\[2ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:\&\&\: e_2 : \mathsf{Bool}}
       & (\textsc{T-And}) \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:||\: e_2 : \mathsf{Bool}}
       & (\textsc{T-Or})
    \end{aligned}
  \]
\end{multicols}

% now the large probabilistic-assign rule full width:
\[
  \frac{
    \Gamma \vdash e_1 : \tau
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_3 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_4 : \tau
  }
  {\Gamma \vdash e_1 \{\,e_2 :\!:\! e_3\}\, e_4 : \tau}
  \quad(\textsc{T-ProbAssign})
\]

\[
  \begin{aligned}
    \frac{\Gamma\vdash e : \mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle
      \quad
      \Gamma\vdash i : \mathsf{Int}_{s,w}}
    {\Gamma\vdash e[i] : \mathsf{Int}_{s,w}}
     & (\textsc{T-Index})     \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Bernoulli}(e_1,e_2) : \mathsf{Int}_{\bot,1}}
     & (\textsc{T-Bernoulli}) \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Uniform}(e_1,e_2) : \mathsf{Int}_{s_2,w_2}}
     & (\textsc{T-Uniform})
  \end{aligned}
\]
\subsection*{Statement Typing Rules}

\[
  \frac{\Gamma\vdash e_1 : \mathsf{Array}<\mathsf{Int}_{s,w}>
  \quad
  \Gamma\vdash i    : \mathsf{Int}_{s,w}
  \quad
  \Gamma\vdash e_2  : \mathsf{Int}_{s,w}}
  {\Gamma\vdash e_1[i] = e_2 \;\mathsf{ok}}
  \quad(\textsc{T-Update})
\]

\begin{multicols}{2}
  % column 1
  \[
    \begin{aligned}
      \frac{\Gamma \vdash e : \mathsf{Bool}
        \quad
        \Gamma \vdash S_1\;\mathsf{ok}
        \quad
        \Gamma \vdash S_2\;\mathsf{ok}}
      {\Gamma \vdash \mathtt{if}(e)\,S_1\,\mathtt{else}\,S_2\;\mathsf{ok}}
       & (\textsc{T-If})       \\[1ex]
      \frac{\Gamma \vdash e : \mathsf{Bool}
        \quad
        \Gamma \vdash S\;\mathsf{ok}}
      {\Gamma \vdash \mathtt{while}(e)\,S\;\mathsf{ok}}
       & (\textsc{T-While})    \\[1ex]
      \frac{\Gamma \vdash S_1\;\mathsf{ok}
        \quad
        \Gamma \vdash S_2\;\mathsf{ok}}
      {\Gamma \vdash \mathtt{try}\,S_1\,\mathtt{catch}\,S_2\;\mathsf{ok}}
       & (\textsc{T-TryCatch})
    \end{aligned}
  \]

  % column 2
  \[
    \frac{\Gamma \vdash e : \mathsf{Bool}}
    {\Gamma \vdash \mathtt{observe}\,(e)\;\mathsf{ok}}
    \quad(\textsc{T-Observe})
  \]

  \[
    \frac{\Gamma \vdash e : \tau_e
      \quad
      x:\tau \in \Gamma}
    {\Gamma \vdash x \;\equiv_{comp}\; e\;\mathsf{ok}}
    \quad(\textsc{T-Assign})
  \]
\end{multicols}

\[
  \tau \equiv_{comp} \tau_e\quad\text{iff}\quad
  \begin{cases}
    \tau_e = \mathsf{Bool},\ \tau = \mathsf{Bool};                                       \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 \le w_2;     \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 - 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 + 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 \le w_2;     \\[4pt]
    \mathbf{arrays\;analogous\;to\;int}
  \end{cases}
\]
